\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-28 paper/algorithm.tex
comments
paper: remove formal proof

\section{Algorithm}
In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}), and we prove that it is optimal, given our search space (\Cref{sec:optimal}). Finally, we show
that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.

@@

Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

@@

\subsection{Optimality}\label{sec:optimal}
We prove the optimality of the final plan selected by showing that our cost-model and algorithm
construct a complete Pareto frontier.

\begin{definition}
$I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
$D$ is the dirty set for a relation, is an abstract imputation operator mapping a set of
attributes with missing values to a set of imputation operations.

For convenience we denote applying the set of imputation operations to a plan $p$ as $I(p)$ and understand that this
is shorthand for $\{ op(p)   | op \in I(\textsc{Dirty}(p)) \}$.
\end{definition}

We can succinctly describe possible imputation transformations to a plan at a given point in the planning algorithm solely through $I$.

\begin{theorem}\label{theorem:no-bad-sit}
By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but 
$\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2') \land p_1' \not \in I(p_2)$
\end{theorem}

\Cref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in~\Cref{fig:bad-sit-diagram}

\begin{figure}
\centering
\begin{tikzpicture}
% before I
\node(p1) {$p_1$};
\node[right=of p1](dom1){$\prec$};
\node[right=of dom1](p2) {$p_2$};

% after I
\node[below=of p1](p1'){$p_1'$};
\node[right=of p1'](dom2) {$\succ$};
\node[right=of dom2](p2'){$p_2'$};

% arrows
\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
\end{tikzpicture}
\caption{A bad Pareto frontier selection would choose $p_2$ over $p_1$, as it is locally optimal, but with the available transformations ($I$) $p_2$  can only reach $p_2'$, which is dominated by the inaccessible $p_1'$.~\Cref{theorem:no-bad-sit} says such a situation cannot arise when constructing Pareto frontiers on a dirty-set basis.}
\label{fig:bad-sit-diagram}
\end{figure}

\begin{proof}
For contradiction, let $p_1$ and $p_2$ be plans such that $\doms{p_2}{p_1}$, $p_1$ is removed from the plan cache
in favor of $p_2$, but there  $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
and $p_1'$ is not reachable from $p_2$ by $I$.

There are two possible cases :

\begin{case}
$\textsc{Dirty}(p_1) = \textsc{Dirty}(p_2)$, in which case $I(p_1) = I(p_2)$, as $I$ uniquely determines the possible transformations. This makes $p_2$ and $p_1$ substitutable, so $p_1'$ is reachable from $p_2$. This is a contradiction.
\end{case}

\begin{case}
$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ relation does not occur within the same dirty set. This is a contradiction
\end{case}

So it must be the case that no such situation arises.
\end{proof}

Given that no desirable plans are discarded by computing Pareto frontiers for each dirty set, the final Pareto frontier (for dirty set $\emptyset$) is optimal.

>>>>

\section{Algorithm}
In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.

@@

\srm{I find this table hard to follow -- I think it'd be good to explain one or two of the rows in detail, and to explain what the set L contains.}
Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

@@

%\subsection{Optimality}\label{sec:optimal}
%We prove the optimality of the final plan selected by showing that our cost-model and algorithm
%construct a complete Pareto frontier.
%
%\begin{definition}
%$I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
%$D$ is the dirty set for a relation, is an abstract imputation operator mapping a set of
%attributes with missing values to a set of imputation operations.
%
%For convenience we denote applying the set of imputation operations to a plan $p$ as $I(p)$ and understand that this
%is shorthand for $\{ op(p)   | op \in I(\textsc{Dirty}(p)) \}$.
%\end{definition}
%
%We can succinctly describe possible imputation transformations to a plan at a given point in the planning algorithm solely through $I$.
%
%\begin{theorem}\label{theorem:no-bad-sit}
%By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but 
%$\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2') \land p_1' \not \in I(p_2)$
%\end{theorem}
%
%\Cref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in~\Cref{fig:bad-sit-diagram}
%
%\begin{figure}
%\centering
%\begin{tikzpicture}
%% before I
%\node(p1) {$p_1$};
%\node[right=of p1](dom1){$\prec$};
%\node[right=of dom1](p2) {$p_2$};
%
%% after I
%\node[below=of p1](p1'){$p_1'$};
%\node[right=of p1'](dom2) {$\succ$};
%\node[right=of dom2](p2'){$p_2'$};
%
%% arrows
%\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
%\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
%\end{tikzpicture}
%\caption{A bad Pareto frontier selection would choose $p_2$ over $p_1$, as it is locally optimal, but with the available transformations ($I$) $p_2$  can only reach $p_2'$, which is dominated by the inaccessible $p_1'$.~\Cref{theorem:no-bad-sit} says such a situation cannot arise when constructing Pareto frontiers on a dirty-set basis.}
%\label{fig:bad-sit-diagram}
%\end{figure}
%
%\begin{proof}
%For contradiction, let $p_1$ and $p_2$ be plans such that $\doms{p_2}{p_1}$, $p_1$ is removed from the plan cache
%in favor of $p_2$, but there  $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
%and $p_1'$ is not reachable from $p_2$ by $I$.
%
%There are two possible cases :
%
%\begin{case}
%$\textsc{Dirty}(p_1) = \textsc{Dirty}(p_2)$, in which case $I(p_1) = I(p_2)$, as $I$ uniquely determines the possible transformations. This makes $p_2$ and $p_1$ substitutable, so $p_1'$ is reachable from $p_2$. This is a contradiction.
%\end{case}
%
%\begin{case}
%$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ relation does not occur within the same dirty set. This is a contradiction
%\end{case}
%
%So it must be the case that no such situation arises.
%\end{proof}
%
%
%Given that no desirable plans are discarded by computing Pareto frontiers for each dirty set, the final Pareto frontier (for dirty set $\emptyset$) is optimal.

### 2017-02-28 paper/introduction.tex
paper: remove formal proof

% Solution
One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we present a new system, called \ProjectName{}, which is the designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression algorithm takes about XXX minutes to train and run
on a 600K row database; prior work~\cite{akande2015empirical} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
\srm{Make numbers above consistent with experiments.}
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
\item \textbf{Proof of optimality:}
 We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

% Solution
One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we present a new system, called \ProjectName{}, which is the designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive -- out experiments show that even a relatively simple regression algorithm algorithms tables about XXX minutes to train and run
on a 600K row database;  prior work~\cite{akande2015empirical} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
\srm{Make numbers above consistent with experiments.}
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}
